Parsing Postfix log files

### 2009-02-27 doc/logparser-architecture.tex
Corrections from proofreading of architecture chapter.

To avoid cluttering the explanation of the parser architecture with the
details involved in parsing Postfix log files and the description of the
parser implemented for this project, the two topics have been separated.
This chapter presents the architecture designed and developed for this
project, beginning with the overall architecture and design, followed by
the three components of the architecture: Framework, Actions, and Rules,
and finishes by describing the characteristics of this architecture.

This chapter centers on the theoretical, implementation-independent aspects
of the architecture; the practical difficulties of writing a parser for
Postfix log files are covered in detail in \sectionref{Postfix Parser
Implementation}.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix vary widely from host to
host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the parser's design aims was to make adding new
rules as easy as possible, to enable administrators to properly parse their
own log files.  To enable this the architecture is divided into three
parts: framework, actions and rules.  Each will be discussed separately,
but first an overview:

@@

    \item [Rules]  The rules are responsible for classifying inputs; they
        specify the action to invoke and the regex that matches the inputs
        and extracts data.  Rules provide an easily extensible method of
        associating inputs with actions.

@@

For each input the framework tries each rule in turn until it finds a rule
that matches the input, then invokes the action specified by that rule.

@@

Decoupling the parsing rules from their associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code, significantly lowering the barrier to entry for casual users
who need to parse new inputs, e.g.\ part-time systems administrators
attempting to combat and reduce spam; it also allows companies to develop
user-extensible parsers without divulging their source code.  Decoupling
the framework, actions, and rules simplifies all three and creates a clear
separation of functionality: the framework provides services to the
actions, but does not need to perform anything specific to the input being
parsed; actions benefit from having services provided by the framework,
freeing them to concentrate on the task of accurately and correctly
processing the inputs and the information provided by rules; rules handle
the low level details of classifying inputs and extracting data from those
inputs, for processing by actions.

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to match the inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Bear in
mind that changes to a parser must be made without adversely affecting
existing parsing, particularly as there may be edge cases that are not
immediately obvious --- \sectionref{yet-more-aborted-delivery-attempts}
describes a complication that occurs only four times in \numberOFlogFILES{}
log files.  Requiring changes to a parser's source code also complicates
upgrades, as the changes must be preserved during the upgrade, and may
clash with changes made by the developer.  This architecture allows the
user to add new rules without editing a parser, unless the new inputs
require cannot be processed by the existing actions.  If the new inputs do
require new functionality, new actions can be added to the parser without
having to modify existing actions; only in the rare event that the new
actions require support from other actions will more extensive changes be
required.

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \acronym{ATN}   & Architecture  & Similarity                  \\
        \tablemiddleline{}%
        Networks        & Parser        & Determines the sequence
                                          of transitions              \\
                        &               & or actions that
                                          constitutes a valid input.  \\
        Transitions     & Actions       & Assemble data and
                                          impose conditions           \\
                        &               & the input must meet to be
                                          accepted as                 \\
                        &               & valid.                      \\
        Abbreviations   & Rules         & Responsible for
                                          classifying input.          \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

The framework takes care of miscellaneous support functions and managing
the parsing process, freeing the programmers writing actions to concentrate
on writing productive code.  It links actions and rules, allowing either to
be improved independently of the other.  It provides shared storage to pass
data between actions, saves and loads state, loads and validates rules,
manages parsing, invokes actions, tracks how often each rule matches to
optimise rule ordering (\sectionref{rule ordering for efficiency}), stores
results of parsing, and miscellaneous other tasks.  Most parsers will
require the same basic functionality from the framework, plus some
specialised support functions XXX EXPAND\@.  The framework is the core of
the architecture and is deliberately quite simple: the rules deal with the
variation in inputs, and the actions deal with the intricacies and
complications encountered when parsing.

The function that finds the rule matching the input and invokes the
requested action can be expressed in pseudo-code (with indentation denoting
flow of control) as:

\begin{verbatim}
for each input:
    for each rule defined by the user:
        if this rule matches the input:
            perform the action specified by the rule
            skip the remaining rules
            process the next input
    warn the user that the input was not parsed
\end{verbatim}

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejections.  There may be many input variants
within one input category; in general each action will handle one category
of inputs, with each rule recognising one input variant.  It is anticipated
that parsers based on this architecture will have a high ratio of rules to
actions, with the aim of having simple rules and clear distinctions between
the inputs parsed by different rules.  An action may process different
input variants in slightly different ways, but large changes in processing
indicate the need for a new action and a new category of input; if an
action becomes unnecessarily complicated is starts to turn into a
monolithic parser, with too much logic contained in a single procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  XXX REWRITE, EXPAND, AND CLARIFY UP
TO THE NEXT XXX\@.  Instead of writing a single monolithic function that
must be modified to support new behaviour, with all the attendant risks of
adversely affecting the existing parser, when a new requirement arises an
independent action can be written to satisfy it.  Sometimes the new action
will require the cooperation of other actions, e.g.\ to set or check a
flag, or save some extra data.  There is a possibility of introducing
failure when modifying existing actions in this way, but the modifications
will be smaller and occur less frequently than with a monolithic
architecture, thus failures will be less likely to occur and will be easier
to test for and diagnose.  XXX FINISH REWRITING\@.  The architecture can be
implemented in an object oriented style, allowing sub-classes to extend or
override actions in addition to adding new actions; because each action is
an independent procedure, the sub-class need only modify the action it is
overriding, rather than reproducing large chunks of functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log \empty{fool check}lines there were
many complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were rectified by later versions of
Postfix, e.g.\ identifying bounce
notifications~\sectionref{identifying-bounce-notifications}); others were
due to the vagaries of process scheduling, client behaviour, and
administrative actions.  All were successfully accommodated in the Postfix
log parser: adding new actions was enough to overcome several of the
complications; others required modifications to a single existing action to
work around the difficulties; the remainder were resolved by adapting
existing actions to cooperate and exchange extra data (via the framework),
changing their behaviour as appropriate based on that extra data.  Every
architecture should aim to make the hard things possible as well as making
the easy things easy; the successful implementation of the Postfix log
parser demonstrates that this architecture achieves that aim.

@@

Rules are responsible for categorising inputs: each rule should recognise
one and only input variant; an input category with multiple input variants
should have multiple rules, one for each variant.  Rules will typically use
a regex when classifying inputs, but other approaches may prove useful for
some applications, e.g.\ comparing fixed strings to the input, or checking
the length of the input; for the remainder of this thesis it will be
assumed that a regex is used.  A rule must specify both the regex to match
against each input and the action to invoke when the match is successful,
but implementations are free to add any other attributes they require;
\sectionref{rule attributes} describes the attributes used in
\parsername{}, and some potential attributes will be discussed later in
this section.

@@

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually adding new
rules, machine learning techniques could be used to automatically generate
new rules.  If this approach was taken the choice of machine learning
technique would be constrained by the size of typical data sets (see
\sectionref{parser efficiency}).  Techniques requiring the full data set
when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser might also dynamically create new rules in response to certain
inputs, e.g.\ diagnostic messages indicating that the program which
produced the input being parsed had read a new configuration file, and had
changed its behaviour appropriately.  These avenues of research and
development have not been pursued by the author, but the architecture
allows them to easily be undertaken independently.

@@

@@

@@

When adding new rules, the rule author must be aware that the new rule may
overlap with an existing rule, i.e.\ some inputs could be parsed by more
than one rule.  The architecture does not try to detect overlapping rules:
that responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the first matching rule wins, and the order in which rules are tried
against each input might change between parser invocations.  Overlapping
rules are frequently a requirement, allowing a more specific rule to match
some inputs and a more general rule to match the remainder, e.g.\
separating \acronym{SMTP} delivery to specific sites from \acronym{SMTP}
delivery to the rest of the world.  Allowing overlapping rules simplifies
both the general rule and the more specific rule; additionally rules from
different sources can be combined with a minimum of prior cooperation or
modification required.  Overlapping rules should have a priority attribute
to specify their relative ordering; negative priorities may be useful for
catchall rules.

XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
PARAGRAPH\@.

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
computation complexity XXX MENTION DEMORGAN''S LAW\@?  COMMENT ON THE
COMPUTATION COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE
SPINNING WHEEL BOOK\@.  Perl 5.10 regexes are more powerful than
traditional regexes: it is possible to match correctly balanced brackets
nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are thus
equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
detecting overlap may require calculating the intersection of two
\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
language cannot always be parsed by a \acronym{CFL}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.
This is certainly not a task to be performed every time the parser runs:
naive detection of overlap amongst the Postfix log parser's
\numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.  When detecting overlap any
conditions attached to rules must be taken into account, because two rules
whose regexes overlap may have conditions attached which prevent the
overlap in practice.  A less naive approach to overlap detection would
first check for overlapping conditions amongst rules, and then check for
overlapping regexes amongst the rules in each group.  Depending on the
conditions employed a rule may fall into more than one group, e.g.\ given
three conditions:

@@

@@

            y & = & cost of checking for overlap between two conditions \\
            x & = & cost of checking for overlap between two regexes    \\
     overlaps & = & set of sets of rules with overlapping conditions    \\

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match every input, and regexes that consume
excessive amounts of CPU time during matching.  Defining a regex to match
all inputs is trivial: \verb!/^/! matches the start of every input.  This
regex would be highlighted by a tool that detects overlapping rules, and
would easily be noticed by visual inspection, but more complex regexes
would be harder to spot.  XXX MAYBE GIVE AN EXAMPLE REGEX\@?  Usually
excessive CPU time is consumed when a regex with a lot of alteration and
variable quantifiers fails to match, but successful matching is generally
quite fast (see~\cite{mastering-regular-expressions} for in-depth
discussion).  XXX THIS NEEDS TO BE EXPANDED, OR MOVED TO FIT IN
ELSEWHERE\@.

    Rules are completely self-contained and can be understood in
        isolation, without reference to any other rules.  This is similar
        to a \acronym{CFG}; in \acronym{CFG} terms the parser rules could
        be described as:

@@

\section{Architecture Characteristics}

\label{Architecture characteristics}

XXX EXPAND ON ALL THESE\@.

\begin{description}

    \item [Matching rules against inputs is simple:]  The first matching
        rule determines the action that will be invoked: there is no
        backtracking to try alternate rules; no attempt is made to pick a
        \textit{best\/} rule.

    \item [Line oriented:]  The architecture is line oriented at present:
        there is no facility for rules to consume more input or push unused
        input back onto the input stream.  The current input can be
        manipulated by actions, using the cascaded parsing feature.  This
        was not a deliberate design decision, rather a consequence of the
        line oriented nature of Postfix log files; more flexible approaches
        could be pursued.

    \item [Context-free rules:]  Rules can not take into account past or
        future inputs when classifying the current input, though conditions
        can affect which rules will be used to classify each input.

    \item [Context-aware actions:] Actions can consult the results (or lack
        of results) of previous actions during execution, providing some
        context sensitivity.

    \item [Cascaded parsing:] Actions can return a modified input to be
        parsed as if read from the input stream, allowing for a simplified
        version of cascaded parsing~\cite{cascaded-parsing}.

    \item [Transduction:]  The architecture can be thought of as
        implementing transduction: it takes data in one form
        and transforms it to another form; \parsername{} transforms log
        files to a database.

    \item [Similarity to Natural Language Processing:] Unlike traditional
        parsers such as those used when compiling a programming language,
        this architecture does not require a fixed grammar specification
        that inputs must adhere to.  The architecture is capable of dealing
        with interleaved inputs, out of order inputs, and ambiguous inputs
        where heuristics must be applied --- all have arisen and been
        successfully accommodated in the Postfix log file parser.

\end{description}

>>>>

To avoid cluttering the explanation of the parser architecture with the
details involved in implementing a parser for Postfix log files, the two
topics have been separated.  This chapter presents the architecture
developed for this project, beginning with the overall architecture and
design, followed by the three components of the architecture: Framework,
Actions, and Rules.  This chapter centers on the theoretical,
implementation-independent aspects of the architecture; the practical
difficulties of writing a parser for Postfix log files are covered in
detail in \sectionref{Postfix Parser Implementation}.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix may vary widely from host
to host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the architecture's design aims was to make adding
new rules to parse new inputs as easy as possible, to enable administrators
to properly parse their own log files.  To enable this the architecture is
divided into three parts: Framework, Actions, and Rules.  Each will be
discussed separately, but first an overview:

@@

    \item [Rules]  The rules are responsible for classifying inputs; each
        rule recognises one input variant, and specifies the action that
        will be invoked to process the input. Rules provide an easily
        extensible method of associating inputs with actions.

@@

For each input the framework tries each rule in turn until it finds a rule
that recognises the input, then invokes the action specified by that rule.
If the input is not recognised by any of the rules, the framework issues a
warning; the framework can continue parsing, although for some inputs
stopping immediately may be preferable.

@@

Decoupling the parsing rules from their associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code, significantly lowering the barrier to entry for casual users
who need to parse new inputs, e.g.\ part-time systems administrators
attempting to combat and reduce spam; it also allows companies to develop
user-extensible parsers without divulging their source code.  Decoupling
the framework, actions, and rules simplifies all three and creates a clear
separation of functionality: the framework manages the parsing process and
provides services to the actions, but does not need to perform anything
specific to the input being parsed; actions benefit from having services
provided by the framework, freeing them to concentrate on the task of
accurately and correctly processing the inputs and the information provided
by rules; rules are responsible for classifying inputs, and extracting data
from those inputs for processing by actions.

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to recognise the inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Bear in
mind that changes to a parser must be made without adversely affecting
existing parsing, particularly as there may be edge cases that are not
immediately obvious --- \sectionref{yet-more-aborted-delivery-attempts}
describes a complication that occurs only four times in \numberOFlogFILES{}
log files.  Requiring changes to a parser's source code also complicates
upgrades of the parser, because the changes must be preserved during the
upgrade, and they may clash with changes made by the developer.  This
architecture allows the user to add new rules without having to edit a
parser, unless the new inputs cannot be processed by the existing actions.
If the new inputs do require new functionality, new actions can be added to
the parser without having to modify existing actions; only in the rare
event that the new actions need to cooperate with existing actions will
more extensive changes be required.

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \acronym{ATN}   & Architecture  & Similarity                  \\
        \tablemiddleline{}%
        Networks        & Parser        & Determines the sequence
                                          of transitions              \\
                        &               & or actions that
                                          constitutes a valid input.  \\
        Transitions     & Actions       & Assemble data and
                                          impose conditions the       \\
                        &               & input must meet to be
                                          accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for
                                          classifying input.          \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

The architecture can be thought of as implementing transduction: it takes
data in one form and transforms it to another form; \parsername{}
transforms log files to a database.  XXX NEED A REFERENCE FOR
TRANSDUCTION\@.

Unlike traditional parsers such as those used when compiling a programming
language, this architecture does not require a fixed grammar specification
that inputs must adhere to.  The architecture is capable of dealing with
interleaved inputs, out of order inputs, and ambiguous inputs where
heuristics must be applied --- all have arisen and been successfully
accommodated in the Postfix log file parser.

@@

The framework manages the parsing process and provides support functions,
freeing the programmers writing actions to concentrate on writing
productive code.  It links actions and rules, allowing either to be
improved independently of the other, and allows new rules to be written
without needing changes to the source code of a parser.  The framework is
the core of the architecture and is deliberately quite simple: the rules
deal with the variation in inputs, and the actions deal with the
intricacies and complications encountered during parsing.

The function that finds the rule matching the input and invokes the
requested action can be expressed in pseudo-code as:

\begin{verbatim}
INPUT:
for each input {
    for each rule defined by the user {
        if this rule matches the input {
            perform the action specified by the rule
            next INPUT
        }
    }
    warn the user that the input was not parsed
}
\end{verbatim}

Most parsers will require the same basic functionality from the framework,
plus some specialised support functions.  XXX EXPAND THIS\@.

XXX TURN THIS INTO A LIST AND EXPAND THE ITEMS\@.  It provides shared
storage to pass data between actions, saves and loads state, loads and
validates rules, manages parsing, invokes actions, stores results of
parsing, and miscellaneous other tasks.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one category of inputs, with each rule recognising one input
variant.  XXX EXPAND AND CLARIFY\@; THE MESSAGE IS HERE, IT JUST NEEDS TO
BE CLEARER\@.  It is anticipated that parsers based on this architecture
will have a high ratio of rules to actions, with the aim of having simple
rules and clear distinctions between the inputs parsed by different rules.
An action may process different input variants in slightly different ways,
but large changes in processing indicate the need for a new action and a
new category of input; if an action becomes unnecessarily complicated it
starts to turn into a monolithic parser, with too much logic contained in a
single procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  XXX REWRITE, EXPAND, AND CLARIFY UP
TO THE NEXT XXX\@.  Instead of writing a single monolithic function that
must be modified to support new behaviour, with all the attendant risks of
adversely affecting the existing parser, when a new requirement arises an
independent action can be written to satisfy it.  Sometimes the new action
will require the cooperation of other actions, e.g.\ to set or check a
flag, or save some extra data.  There is a possibility of introducing
failure when modifying existing actions in this way, but the modifications
will be smaller and occur less frequently than with a monolithic
architecture, thus failures will be less likely to occur and will be easier
to test for and diagnose.  XXX FINISH REWRITING\@.  The architecture can be
implemented in an object oriented style, allowing sub-classes to extend or
override actions in addition to adding new actions; because each action is
an independent procedure, the sub-class need only modify the action it is
overriding, and can inherit the others rather than reproducing large chunks
of functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log~lines there were
many complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were rectified by later versions of
Postfix, e.g.\ identifying bounce
notifications~\sectionref{identifying-bounce-notifications}); others were
due to the vagaries of process scheduling, client behaviour, and
administrative actions.  All were successfully accommodated in the Postfix
log parser: adding new actions was enough to overcome several of the
complications; others required modifications to a single existing action to
work around the difficulties; the remainder were resolved by adapting
existing actions to cooperate and exchange extra data (via the framework),
changing their behaviour as appropriate based on that extra data.  Every
architecture should aim to make the hard things possible as well as making
the easy things easy; the successful implementation of the Postfix log
parser demonstrates that this architecture achieves that aim.

Actions are not context-sensitive, but they can consult the results (or
lack of results) of previous actions during execution, providing some
context sensitivity.  XXX EXPAND\@.

@@

Rules are responsible for categorising inputs: each rule should recognise
one and only input variant.  An input category with multiple input variants
should have multiple rules, one for each variant.  Rules will typically use
a regex when classifying inputs, but other approaches may prove useful for
some applications, e.g.\ comparing fixed strings to the input, or checking
the length of the input; for the remainder of this thesis it will be
assumed that a regex is used.  Each rule must specify at least the regex to
recognise inputs and the action to invoke when classification is
successful, but implementations are free to add any other attributes they
require; \sectionref{rule attributes} describes the attributes used in
\parsername{}, and some potential attributes will be discussed later in
this section (\sectionref{attaching conditions to rules} and
\sectionref{prioritisation and rule ordering}).  Classifying inputs using
the rules is simple: the first rule to recognise the input determines the
action that will be invoked: there is no backtracking to try alternate
rules; no attempt is made to pick a \textit{best\/} rule.

@@

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually editing
existing rules or adding new rules, machine learning techniques could be
used to automate the process.  If this approach was taken the choice of
machine learning technique would be constrained by the size of typical data
sets (see \sectionref{parser efficiency}).  Techniques requiring the full
data set when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser could also dynamically create new rules in response to certain
inputs, e.g.\ XXX GIVE A GOOD EXAMPLE\@.  These avenues of research and
development have not been pursued by the author, but the architecture
allows them to easily be undertaken independently.

@@

\label{attaching conditions to rules}

@@

\label{prioritisation and rule ordering}

@@

\label{overlapping rules in architecture}

XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?

When adding new rules, the rule author must be aware that the new rule may
overlap with one or more existing rules, i.e.\ some inputs could be parsed
by more than one rule.  The architecture does not try to detect overlapping
rules: that responsibility is left to the author of the rules.
Unintentionally overlapping rules lead to inconsistent parsing and data
extraction because the first matching rule wins, and the order in which
rules are tried against each input might change between parser invocations.
Overlapping rules are frequently a requirement, allowing a more specific
rule to match some inputs and a more general rule to match the remainder,
e.g.\ separating \acronym{SMTP} delivery to specific sites from
\acronym{SMTP} delivery to the rest of the world.  Allowing overlapping
rules simplifies both the general rule and the more specific rule;
additionally rules from different sources can be combined with a minimum of
prior cooperation or modification required.  Overlapping rules should have
a priority attribute to specify their relative ordering; negative
priorities may be useful for catchall rules.

XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
PARAGRAPH\@.  XXX DO I NEED TO MENTION CFG/CFL AT ALL\@?

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
BOOK\@.  Perl 5.10 regexes are more powerful than traditional regexes: it
is possible to match correctly balanced brackets nested to an arbitrary
depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
detecting overlap may require calculating the intersection of two
\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
language cannot always be parsed by a \acronym{CFL}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
is certainly not a task to be performed every time the parser runs: naive
detection of overlap amongst the Postfix log parser's \numberOFrules{}
rules would require calculating \numberOFruleINTERSECTIONS{} intersections.
When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the overlap in practice.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlapping regexes amongst the rules in each set
of rules with overlapping conditions.  Depending on the conditions employed
a rule may fall into more than one group, e.g.\ given these three
conditions:

@@

    % Reduce inter-item spacing because each item is a single line.
    \setlength{\itemsep}{0pt}

@@

            $y$ & = & cost of checking for overlap between two conditions \\
            $x$ & = & cost of checking for overlap between two regexes    \\
     $overlaps$ & = & set of sets of rules with overlapping conditions    \\

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be highlighted by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Usually excessive CPU time
is consumed when a regex with a lot of alteration and variable quantifiers
fails to match, but successful matching is generally quite fast
(see~\cite{mastering-regular-expressions} for in-depth discussion).  XXX
THIS NEEDS TO BE EXPANDED, AND GIVE AN EXAMPLE\@.

XXX PATHOLOGICAL CONDITIONS\@.

\subsection{XXX TO BE DEALT WITH}

XXX DUNNO WHERE TO PUT THIS\@; IT SHOULD PROBABLY GO IN THE RULES SECTION,
BUT WHERE\@?

Rules are self-contained and can be understood in isolation, without
reference to other rules.  This is similar to a \acronym{CFG}; in
\acronym{CFG} terms the parser rules could be described as:

@@

Rules can not take into account past or future inputs when classifying the
current input, though conditions can affect which rules will be used to
classify each input.  The architecture is line oriented at present: there
is no facility for rules to consume more input or push unused input back
onto the input stream.  The current input can be manipulated by actions,
using the cascaded parsing feature.  This was not a deliberate design
decision, rather a consequence of the line oriented nature of Postfix log
files; more flexible approaches could be pursued.
